---
title: "`r paste(params$sample_location, '- Soil Profile Textural Analysis')`"
author: "`r params$author`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  sample_location: "Site"
  author: ""
  data_files: NULL  # List of dataset configs: list(path, label, colour)
  min_replicates: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

library(tidyverse)
library(scales)
library(janitor)
library(knitr)
library(ggtern)
library(ggsoiltexture)
library(soiltexture)

# ╔══════════════════════════════════════════════════════════════╗
# ║                        CONFIG                               ║
# ╠══════════════════════════════════════════════════════════════╣
# ║  Add one list() entry per dataset.                          ║
# ║  label  – display name in all plots / tables                ║
# ║  colour – hex colour for this dataset                       ║
# ║                                                             ║
# ║  SAMPLE NAMING CONVENTION                                   ║
# ║  <sample_id>_<meta_code>_<rep_number>_<optional_note>       
# ║  ensure consistency in depth unites ie. meters vs centimeters 
# ║                                                             ║
# ║  Examples:                                                  ║
# ║    25-CR-R-0.4_B_2                                          ║
# ║      sample_id  = 25-CR-R-0.4                               ║
# ║      meta_code  = B   (treatment, prep method, etc.)        ║
# ║      rep        = 2                                         ║
# ║      note       = (none)                                    ║
# ║                                                             ║
# ║    25-CR-R-0.4_B_2_sonic                                    ║
# ║      sample_id  = 25-CR-R-0.4                               ║
# ║      meta_code  = B                                         ║
# ║      rep        = 2                                         ║
# ║      note       = sonic   (free text, shown in tables)      ║
# ║                                                             ║
# ║  Samples sharing the same sample_id + meta_code within a   ║
# ║  dataset are treated as replicates of each other.           ║
# ╚══════════════════════════════════════════════════════════════╝

# Validate required parameters
if (is.null(params$data_files)) {
  stop("Error: data_files parameter is required.
       Provide a list of datasets with 'path', 'label', and 'colour' fields.
       
       Example:
       data_files = list(
         list(path = 'data/dataset1.txt', label = 'Dataset 1', colour = '#4E79A7'),
         list(path = 'data/dataset2.txt', label = 'Dataset 2', colour = '#F28E2B')
       )")
}

# Validate that all files exist
for (dataset in params$data_files) {
  if (!file.exists(dataset$path)) {
    stop(paste("Error: File not found:", dataset$path))
  }
}

data_files <- params$data_files
min_replicates <- params$min_replicates

# Minimum replicates required to include a sample group
# Override at render time: rmarkdown::render(..., params = list(min_replicates = 3))
min_replicates <- params$min_replicates

# ── END CONFIG ───────────────────────────────────────────────────────────────
```

```{r load-and-parse, include=FALSE}

# Dataset colour lookup
dataset_colours <- setNames(
  map_chr(data_files, "colour"),
  map_chr(data_files, "label")
)

# Defines a function to read one instrument export 
read_instrument_file <- function(path, dataset_label, ref_bin_names = NULL) {
  dat <- readr::read_delim(
    file           = path,
    delim          = "\t",
    locale         = readr::locale(encoding = "UTF-16LE"),
    show_col_types = FALSE
  ) |>
    dplyr::mutate(dplyr::across(
      dplyr::where(is.character),
      ~ gsub("\\?m", "um", .x)
    )) |>
    janitor::clean_names(replace = c("\u00B5" = "u", "\u03bc" = "u")) |>
    mutate(
      sample_name = sample_name |>
        str_remove("^Average of\\s*'") |>
        str_remove("'\\s*$"),
      dataset = dataset_label
    )
  
  # If reference bin names provided, rename this file's bin columns to match
  if (!is.null(ref_bin_names)) {
    bin_cols <- names(dat)[str_detect(names(dat), "^x\\d")]
    if (length(bin_cols) == length(ref_bin_names)) {
      dat <- dat |> rename(setNames(bin_cols, ref_bin_names))
    } else {
      warning(sprintf("File %s has %d bin columns, reference has %d — skipping rename",
                      path, length(bin_cols), length(ref_bin_names)))
    }
  }
  dat
}

# Load all datasets into a single 

# Read first file to get reference bin names
ref_file     <- data_files[[1]]
ref_dat      <- read_instrument_file(ref_file$path, ref_file$label)
ref_bin_names <- names(ref_dat)[str_detect(names(ref_dat), "^x\\d")]

# Load all datasets using reference bin names
all_raw <- map_dfr(data_files, function(f) {
  if (!file.exists(f$path)) {
    warning("File not found, skipping: ", f$path)
    return(NULL)
  }
  read_instrument_file(f$path, f$label, ref_bin_names = ref_bin_names)
})

# ── Split binned grain size data from vs summary stats 
id_cols    <- c("sample_name", "dataset")
is_bin_col <- str_detect(names(all_raw), "^x\\d")

## psd_long is the binned data
psd_long <- all_raw[, c(id_cols, names(all_raw)[is_bin_col]), drop = FALSE] |>
  pivot_longer(
    cols      = matches("^x\\d"),
    names_to  = "grain_um",
    values_to = "perc_vol_in"
  ) |>
  mutate(
    grain_um = as.numeric(
      str_replace_all(str_remove(grain_um, "^x"), "_", "."))
  )
## summ_long is the summary data from Malvern's calulations
summ_long <- all_raw |>
  select(-matches("^x\\d")) |>
  pivot_longer(
    cols      = where(is.numeric),
    names_to  = "metric",
    values_to = "value_raw"
  )

# ── Parse sample names ────────────────────────────────────────────────────────
## defines a function to read out my sample name code
## Sample codes is samplename-depth_treatmentcode_replicatenumber_notes

parse_sample_name <- function(nm) {
  tibble(sample_name = nm) |>
    mutate(
      # look to see if there is any text at the end of the name with special notes
      note      = str_extract(sample_name, "(?<=_\\d{1,3}_)[^_].+$"),
      # identifies the replicate number
      rep_num   = str_extract(sample_name, "\\d+(?=(_[^_]+)?$)"),
      # identifies the pretreatment code
      meta_code = str_extract(sample_name,
                               paste0("(?<=_)[A-Z](?=_", rep_num, ")")),
      # identifies the base sample name including the depth value 
      sample_id = str_remove(
        sample_name,
        paste0("_", replace_na(meta_code, ""),
               if_else(!is.na(meta_code), "_", ""),
               rep_num,
               if_else(!is.na(note), paste0("_", note), ""),
               "$")
      ),
      # extracts the depth value from the base name 
      depth_val = as.numeric(str_extract(sample_id, "[\\d.]+$"))
    )
}
# generates a table that summarizes the details from the sample name 
# key table that everything else joins against to get sample_id, depth_val, meta_code, and note
sample_info <- psd_long |>
  distinct(sample_name, dataset) |>
  bind_cols(
    parse_sample_name(
      psd_long |> distinct(sample_name, dataset) |> pull(sample_name)
    ) |> select(-sample_name)
  )

# ── Replicate groups ──────────────────────────────────────────────────────────
## one row per unique sample group, with everything needed to loop through them and pull their data.
replicate_groups <- sample_info |>
  group_by(dataset, sample_id, depth_val, meta_code) |>
  summarise(
    n_replicates = n(),
    sample_names = list(sample_name),
    notes        = list(unique(na.omit(note))),
    .groups      = "drop"
  ) |>
  filter(n_replicates >= min_replicates) |>
  mutate(dataset = factor(dataset, levels = map_chr(data_files, "label"))) |>
  arrange(dataset, depth_val) |>
  mutate(dataset = as.character(dataset))

# ── Attach parsed fields back to psd_long and summ_long ──────────────────────
psd_long <- psd_long |>
  left_join(
    sample_info |> select(sample_name, dataset, sample_id, depth_val, meta_code, note),
    by = c("sample_name", "dataset")
  )

summ_long <- summ_long |>
  left_join(
    sample_info |> select(sample_name, dataset, sample_id, depth_val, meta_code, note),
    by = c("sample_name", "dataset")
  )

# ── Pre-compute CV% for all groups in one pass ────────────────────────────────
# D10/D50/D90 read from instrument export columns — not derived from bin data.
cv_all_raw <- summ_long |>
  filter(metric %in% c("dx_10", "dx_50", "dx_90")) |>
  mutate(value = as.numeric(value_raw)) |>
  # grab only samples from the same dataset that belong to the same replicate group
  inner_join(
    replicate_groups |> select(dataset, sample_id, meta_code),
    by = c("dataset", "sample_id", "meta_code")
  ) |>
  #calculates mean, sd, and coefficient of variation for each metric group d10, d50, d90 for each replicate group
  group_by(dataset, sample_id, depth_val, meta_code, metric) |>
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    sd_val   = sd(value,   na.rm = TRUE),
    cv_pct   = (sd_val / mean_val) * 100,
    .groups  = "drop"
  ) |>
  # assigns a qualifier for each to indicate how well we replicated samples. 
  mutate(
    metric_label = recode(metric,
      "dx_10" = "D10", "dx_50" = "D50", "dx_90" = "D90"),
    status = case_when(
      cv_pct < 3  ~ "Excellent",
      cv_pct < 5  ~ "Good",
      cv_pct < 10 ~ "Acceptable",
      TRUE        ~ "Poor"
    )
  )

# ── Pre-split psd_long by replicate group for fast loop access ────────────────
psd_split <- psd_long |>
  filter(sample_id %in% replicate_groups$sample_id) |>
  group_by(dataset, sample_id, meta_code) |>
  group_split() |>
  setNames(
    psd_long |>
      filter(sample_id %in% replicate_groups$sample_id) |>
      distinct(dataset, sample_id, meta_code) |>
      arrange(dataset, sample_id, meta_code) |>
      mutate(key = paste(dataset, sample_id, meta_code, sep = "||")) |>
      pull(key)
  )
```

---

# Overview

```{r overview, echo=FALSE}
# text showing number of datasets loaded 
cat(sprintf("Datasets loaded:      %d\n", n_distinct(replicate_groups$dataset)))

# text showing total number of sample groups with the minimum number of replicates required 
cat(sprintf("Sample groups found:  %d  (minimum %d replicates each)\n\n",
            nrow(replicate_groups), min_replicates))
# displays the replicate groups in a summary table 
replicate_groups |>
  mutate(
    Notes      = map_chr(notes, ~ if (length(.x) == 0) "—" else paste(.x, collapse = ", ")),
    Replicates = map_chr(sample_names, ~ paste(.x, collapse = ", "))
  ) |>
  select(
    Dataset     = dataset,
    `Sample ID` = sample_id,
    Depth       = depth_val,
    Code        = meta_code,
    Notes,
    N           = n_replicates,
    Replicates
  ) |>
  # kable() is used to generate a table in HTML format 
  kable()
```
---


# USDA Soil Texture Triangle

Soil textural triangle showing each dataset plotted as different colors. 

```{r texture-triangle, echo=FALSE, fig.height=9, fig.width=10, message=FALSE, warning=FALSE}

replicate_texture <- psd_long |>
  filter(sample_id %in% replicate_groups$sample_id) |>
  mutate(
    fraction = case_when(
      grain_um < 2                     ~ "clay",
      grain_um >= 2  & grain_um < 50   ~ "silt",
      grain_um >= 50 & grain_um < 2000 ~ "sand",
      TRUE                             ~ "gravel"
    )
  ) |>
  group_by(dataset, sample_name, sample_id, depth_val, fraction) |>
  summarise(perc = sum(perc_vol_in, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = fraction, values_from = perc, values_fill = 0) |>
  mutate(
    total     = clay + silt + sand,
    clay_norm = 100 * clay / total,
    silt_norm = 100 * silt / total,
    sand_norm = 100 * sand / total
  ) |>
  rename(SAND = sand_norm, CLAY = clay_norm, SILT = silt_norm) |> 
  filter(total > 0) |>
  mutate(dataset = factor(dataset, levels = map_chr(data_files, "label")))

## Plot data on USDA soil textural triangle 

### Using soiltexture package 

# TT.plot expects a data frame with columns named CLAY, SILT, SAND in 0-100
# tt_data <- replicate_texture |>
#   transmute(
#     CLAY  = clay_norm,
#     SILT  = silt_norm,
#     SAND  = sand_norm,
#     dataset   = dataset,
#     depth_val = depth_val
#   ) |>
#   as.data.frame()  # TT.plot doesn't like tibbles
# 
# # Draw the triangle with USDA classification
# TT.plot(
#   class.sys   = "USDA.TT",
#   tri.data    = tt_data,
#   col         = dataset_colours[tt_data$dataset],
#   cex         = 1 + (tt_data$depth_val / max(tt_data$depth_val)) * 2,  # size by depth
#   pch         = 16,
#   main        = "USDA Soil Texture — All Datasets"
# )
# 
# # Add a legend manually
# legend("topleft", 
#        legend = names(dataset_colours),
#        col    = dataset_colours,
#        pch    = 16,
#        title  = "Dataset")

### Using ggsoiltexture package 
# p_texture <- ggsoiltexture(replicate_texture,
#                             show_grid = FALSE,
#                             class     = "USDA") +
#   geom_point(aes(color = dataset,
#                  alpha = depth_val),
#              size = 3) +
#   scale_colour_manual(values = dataset_colours) +
#   scale_alpha_continuous(name = "Depth (m)", range = c(0.2, 1)) +
#   labs(
#     title  = "USDA Soil Texture — All Datasets",
#     T      = "Clay (%)",
#     L      = "Silt (%)",
#     R      = "Sand (%)",
#     colour = "Dataset"
#   ) +
#   theme_bw() +
#   theme_showarrows() +
#   theme(legend.position = "bottom")
# 
# print(p_texture)

### Using ggtern package 

# Build USDA class polygons from soiltexture vertex data
vertices <- TT.vertices.tbl(class.sys = "USDA.TT")
classes  <- TT.classes.tbl(class.sys  = "USDA.TT")

# Convert point index strings to polygon data frames
usda_polygons <- purrr::map_dfr(seq_len(nrow(classes)), function(i) {
  pts <- as.integer(str_split(classes[i, "points"], ",\\s*")[[1]])
  vertices[match(pts, vertices$points), ] |>
    mutate(class = classes[i, "abbr"])
})

# Centroid of each class for labels
usda_labels <- tribble(
  ~class,    ~CLAY,  ~SILT,  ~SAND,
  "Clay",    0.700,  0.150,  0.150,
  "SiClay",  0.500,  0.450,  0.050,
  "SaClay",  0.450,  0.050,  0.500,
  "ClLoam",  0.320,  0.340,  0.340,
  "SiClLo",  0.330,  0.570,  0.100,
  "SaClLo",  0.260,  0.120,  0.620,
  "Loam",    0.180,  0.410,  0.410,
  "SiLoam",  0.120,  0.650,  0.230,
  "SaLoam",  0.080,  0.250,  0.670,
  "Silt",    0.060,  0.880,  0.060,
  "Lo Sand", 0.060,  0.080,  0.860,
  "Sand",    0.020,  0.050,  0.930
)

# Plot
p_tern <- ggtern(data = replicate_texture,
       aes(x = SAND, y = CLAY, z = SILT)) +
  geom_polygon(data     = usda_polygons,
               aes(x = SAND, y = CLAY, z = SILT, group = class),
               fill     = NA,
               colour   = "grey60",
               linewidth = 0.3) +
  annotate("text",
         x      = usda_labels$SAND,
         y      = usda_labels$CLAY,
         z      = usda_labels$SILT,
         label  = usda_labels$class,
         size   = 5,
         colour = "grey40")+
  geom_point(aes(colour = dataset, alpha = depth_val),
             size = 3) +
  scale_colour_manual(values = dataset_colours) +
  scale_alpha_continuous(name  = "Depth (m)", range = c(0.2, 1), guide = "none") +
  labs(
    title  = "USDA Soil Texture — All Datasets",
    subtitle = "Point transparency decreases with depth (shallow = faint, deep = solid)",
    T      = "Clay (%)",
    R      = "Silt (%)",
    L      = "Sand (%)",
    colour = "Dataset"
  ) +
  theme_bw() +
  theme_clockwise() +
  theme_showarrows() +
  theme(legend.position = "bottom")

print(p_tern)
```
---

# Depth Comparisons

*Each section shows the grain size distributions at one sample depth,
with each dataset plotted as a separate line and showing all replicates.*

```{r depth-comparison, results='asis', echo=FALSE, fig.height=5, fig.width=9}

depths_present <- replicate_groups |>
  distinct(sample_id, depth_val) |>
  arrange(depth_val)

for (i in seq_len(nrow(depths_present))) {

  sid   <- depths_present$sample_id[i]
  depth <- depths_present$depth_val[i]

  cat(sprintf("\n\n## %s\n\n", sid))

  depth_psd <- psd_long |>
    filter(sample_id == sid) |>
    group_by(dataset, sample_name, grain_um) |>
    summarise(mean_vol = mean(perc_vol_in, na.rm = TRUE), .groups = "drop") |>
    mutate(dataset = factor(dataset, levels = map_chr(data_files, "label")))

  p <- depth_psd |>
    ggplot(aes(x = grain_um, y = mean_vol,
               colour = dataset, group = sample_name)) +
    geom_line(linewidth = 1, alpha = 0.8) +
    scale_x_log10(labels = label_number()) +
    scale_y_continuous(limits = c(0, NA)) +
    scale_colour_manual(values = dataset_colours) +
    labs(
      title    = paste("Grain size distributions —", sid),
      subtitle = "Mean of replicates per dataset",
      x        = "Grain size (µm)",
      y        = "Volume %",
      colour   = "Dataset"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")

  print(p)
  cat("\n\n---\n\n")
}
```

---

# Per-Sample Distributions

*Each section shows the replicate curves for one sample plus a D10/D50/D90 coefficient of variation (CV%) table to assess sampling replicability*

```{r per-sample, results='asis', echo=FALSE, fig.height=5, fig.width=9}

for (i in seq_len(nrow(replicate_groups))) {

  row    <- replicate_groups[i, ]
  snames <- row$sample_names[[1]]
  sid    <- row$sample_id
  ds     <- row$dataset
  meta   <- row$meta_code
  notes  <- unlist(row$notes)

  meta_str <- if (!is.na(meta)) paste0(" [", meta, "]") else ""
  note_str <- if (length(notes) > 0) paste0(" {", paste(notes, collapse = ", "), "}") else ""
  lbl      <- paste0(sid, meta_str, note_str)

  # Use pre-split data
  key <- paste(ds, sid, meta, sep = "||")
  group_psd <- psd_split[[key]]

  # Section heading
  cat(sprintf("\n\n## %s  <small>(%s)</small>\n\n", lbl, ds))

  if (length(notes) > 0) {
    cat(sprintf("> **Notes:** %s\n\n", paste(notes, collapse = ", ")))
  }

  # Grain size distribution plots
  p <- group_psd |>
    ggplot(aes(x = grain_um, y = perc_vol_in, color = sample_name)) +
    geom_line(linewidth = 1, alpha = 0.7) +
    scale_x_log10(labels = label_number()) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(
      title    = lbl,
      subtitle = paste0(ds,
                        if (length(notes) > 0) paste0("  |  ", paste(notes, collapse = ", ")) else ""),
      x        = "Grain size (µm)",
      y        = "Volume %",
      colour   = "Replicate"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")

  print(p)
  cat("\n\n")

  # CV% table — pulled from pre-computed cv_all_raw
  cv_tbl <- cv_all_raw |>
    filter(dataset == ds, sample_id == sid,
           if (!is.na(meta)) meta_code == meta else is.na(meta_code))

  cat("**D10 / D50 / D90 replicate CV%**\n\n")

  cv_tbl |>
    select(
      Percentile  = metric_label,
      `Mean (µm)` = mean_val,
      `SD (µm)`   = sd_val,
      `CV%`       = cv_pct,
      Status      = status
    ) |>
    mutate(across(where(is.numeric), ~ round(.x, 2))) |>
    kable() |>
    print()

  cat("\n\n---\n\n")
}
```


---

**Report generated:** `r Sys.time()`
